How fluid-mechanical erosion creates anisotropic porous media

### 2022-06-29 docs/PhysicsPaper/StokesErosion.tex
NM

The previous section shows intriguing features to arise from the erosion of a random initialization of solid bodies. The observations natural raise the question: are the trends specific to the particular simulation shown or do these features emerge robustly across different simulations? To answer this question, we now analyze the same quantities---permeability, tortuosity, and anisotropy---over an ensemble of many simulations having different numbers of bodies and different random initializations. We analyze runs having $M$ = 20, 40, 60, 80, and 100 bodies, with at least three different instances of each. This data set represents roughly 2,100 hours of computational time.

@@

Figure~\ref{fig5} shows the permeability and anisotropy measurements for the complete data set. In particular, for a given number of initial bodies ($M$ = 20, 40, 60, 80, or 100), the mean of each quantity (permeability or anisotropy) is shown with the vertical error bars representing the standard deviation. Different initial numbers of bodies, $M$, are shown by different colors. As seen in Figures~\ref{fig5}(a)--(b), the trends of increasing permeabilities are well grouped by the number of initial bodies. That is, permeability monotonically decreases with the number of bodies and the error bars shows little overlap between different values of $M$ until very late in the simulation when the bodies have nearly vanished.

Figure~\ref{fig5}(c) shows the corresponding statistics of anisotropy, $\anis$. Here, the anisotropy is not as well separated by $M$. That is, given the error-bars, there is significant overlap in the signal of $\anis$ across different values of $M$. One notable feature is a second rise in $\anis$ that occurs late in the simulation only for the larger values of $M$ = 60, 80, and 100. As before, the anisotropy can be decomposed into configuration and shape components as shown in Figures~\ref{fig5}(d)--(e). Here, we see the late surge in configurational anisotropy for these larger values of $M = $ 60, 80, 100. For these runs with a large number of bodies, the latest stages of erosion are dominated by channelization which substantially promotes the configurational anisotropy. The shape anisotropy, meanwhile, takes a more regular, parabolic form that is relatively insensitive to the number of bodies.

@@

%(a)--(b) The longitudinal (a) and transverse (b) tortuosity measurements are well grouped with respect to the number of bodies.

Similar statistical analysis is performed on the tortuosity measurements and shown in Figure~\ref{fig6}(a)--(b). In all simulations, the longitudinal tortuosity decreases with time as erosion allows passive tracers to take more direct paths when traveling in the flow direction. In contrast, Fig.~\ref{fig6}(b) shows the transverse tortuosity can increase with time, as vertically-traveling passive tracers must work to circumvent horizontally aligned bodies carved by erosion.

The tortuosity ratio, $(T_2 -1)/(T_1 -1)$, provides a second proxy for medium anisotropy to complement the permeability measurements. Figure~\ref{fig6}(c) shows that the tortuosity ratio behaves similarly across different values of $M$. The ratio increases substantially with erosion, reaching a peak of roughly $(T_2 -1)/(T_1 -1) = 8$, before descending during the final stages of erosion. As before the tortuosity ratio can be decomposed into configurational and shape components, as shown in Figures~\ref{fig6}(d)--(e). Here, we see that the shape component is the main contribution, and the configurational component of the tortuosity ratio is minimal. This observation fits with the intuition that the main hindrance to a passive tracer should be the shape of individual obstructions, rather than how those obstructions are arranged relative to one another. As such, the tortuosity ratio is a less sensitive indicator of channelization than is the permeability ratio.

>>>>

The previous section shows intriguing features to arise from the erosion of a random initialization of solid bodies. The observations natural raise the question: are the trends specific to the particular simulation shown or do these features emerge robustly across different initial conditions? To answer this question, we now analyze the same quantities---permeability, tortuosity, and anisotropy---over an ensemble of simulations having different numbers of bodies and different random initializations. We analyze runs having $M$ = 20, 40, 60, 80, and 100 bodies, with at least three different instances of each. This data set represents roughly 2,100 hours of computational time.

@@

Figure~\ref{fig5} shows the permeability and anisotropy measurements for the complete data set. In particular, for a given number of initial bodies ($M$ = 20, 40, 60, 80, or 100), the mean of each quantity (permeability or anisotropy) is plotted against increasing porosity, with the standard deviation across the runs shown by the vertical error bars. Different initial numbers of bodies, $M$, are shown by different colors. As seen in Figures~\ref{fig5}(a)--(b), the trends of increasing permeabilities are well grouped by the number of initial bodies. That is, permeability monotonically decreases with the number of bodies and the error bars shows little overlap between different values of $M$ until very late in the simulation when the bodies have nearly vanished.

	Figure~\ref{fig5}(c) shows the corresponding statistics of anisotropy, $\anis$. Notice that the anisotropy is not as well separated by $M$. That is, given the error-bars, there is significant overlap in the signal of $\anis$ across different values of $M$. For all values of $M$, the anisotropy initially increases and reaches a peak of roughly $\anis \approx 4$. Interestingly, for the runs featuring a large number of bodies ($M$ = 60, 80, and 100) there is a second rise in anisotropy that occurs late in the simulation and results in a higher peak of roughly $\anis \approx 6$. Thus, the late surge of anisotropy, first observed in Fig.~\ref{fig3}, is not specific to that particular simulation. Rather, it occurs robustly across all simulations as long as the initial number of bodies is sufficiently large, $M \ge 60$.

 	As before, the anisotropy can be decomposed into configuration and shape components, $\anis_C$ and $\anis_S$ respectively, and the statistics of these quantities are shown in Figs.~\ref{fig5}(d)--(e). The shape anisotropy, seen in Fig.~\ref{fig5}(e), takes a fairly regular, parabolic arc --- first increasing due to the shapes carved by erosion and then decreasing as the bodies vanish. This behavior is consistent across all of the values of $M$. The configurational anisotropy (Fig.~\ref{fig5}(d)), however, shows less regular behavior. For runs with a smaller number of bodies ($M$ = 20 and 40) the configurational anisotropy remains relatively small throughout the entire simulation, indicating that the majority of observed anisotropy is due to the individual shapes of bodies. For the runs with a higher body count ($M $ = 60, 80, and 100), though, the configurational anisotropy grows moderately and then surges late in the simulation, as was observed in the single 80-body simulation from Fig.~\ref{fig3}. Since the shape anisotropy is decreasing at this time, the observed second rise in the total anisotropy is due entirely to this surge in configurational component. For these high-body count runs, the latest stages of erosion are dominated by channelization, which substantially promotes the configurational anisotropy over shape anisotropy. That is, the large-scale arrangement of the bodies has greater effect than the shape of individual bodies at this late stage.

@@

We show in Figs.~\ref{fig6}(a)--(b) similar statistical analysis performed on the longitudinal and transverse tortuosity measurements, $T_1$ and $T_2$. Figure~\ref{fig6}(a) shows that, in all simulations, the longitudinal tortuosity decreases with time as erosion allows passive tracers to take more direct paths when traveling in the flow direction. In contrast, Fig.~\ref{fig6}(b) shows the transverse tortuosity increases with time, as vertically-traveling passive tracers must work to circumvent the horizontally aligned bodies carved by erosion.

As before, the tortuosity ratio, $(T_2 -1)/(T_1 -1)$, provides a second proxy for medium anisotropy to complement the permeability-based quantity. Figure~\ref{fig6}(c) shows that the tortuosity ratio behaves similarly across all simulations and all different values of $M$. The ratio increases substantially with erosion, reaching a peak of roughly $(T_2 -1)/(T_1 -1) \approx 8$, before descending during the final stages of erosion. As before the tortuosity ratio can be decomposed into configurational and shape components, as shown in Figures~\ref{fig6}(d)--(e). Here, we see that the shape component is the main contribution and the configurational component remains minimal. This observation fits with the intuition that the main hindrance to a passive tracer should be the shape of individual obstructions, rather than how those obstructions are arranged relative to one another. As such, the tortuosity ratio is a less sensitive indicator of channelization than is the permeability ratio.
